---
title: "GR Crash Data Exploration"
author: "Luke Cadagin"
date: "11/1/2021"
output: html_document
---

```{r loading packages, message = FALSE}
library(tidyverse)
library(lubridate)
library(sf)
library(osmdata)
library(here)
library(tmap)
library(tmaptools)
library(ggforce)
```

```{r}
crash_data <- read_csv(here::here("data", "CGR_Crash_Data.csv"))
head(crash_data)
```



```{r}
glimpse(crash_data)
```


The below speedlimits appear to be incorrect.  If a speedlimit is N/A or Unknown, it appears it is typically marked with 99.  No Michigan road had a speelimit over 70 unitl 2017.

```{r}
crash_data %>% 
  filter(SPEEDLIMIT > 70, SPEEDLIMIT != 99) %>% 
  select(SPEEDLIMIT, PRNAME, INTERNAME, YEAR)
```

Note:

REFDIR is a variable indicating the direction from the impact to the intersecting road.
REFDIST is a variable indicating the distance from the impact to the intersecting road.
Note that this can be in feet of miles, but the dataset failed to include this field.

UD10NUM is a variable indicating the Crash Report "UD-10" Identification number.  This is useful for matching a specific report to a record.

PRNAME is the primary road name.
PRNO seems to be a special numerical designation for Grand Rapids roads used by the police

```{r}
crash_data %>% 
  filter(YEAR == 2008) %>%
  distinct(PRNO)
  #select(PRNO) #%>% 
  #ggplot() +
  #geom_histogram(mapping = aes(x = PRNO, color = PRNAME))
```

```{r}
crash_data %>% 
  filter(YEAR == 2008, DRIVER1AGE <150) %>%
  ggplot() +
  geom_histogram(mapping = aes(x = DRIVER1AGE))
```

```{r}
head(crash_data)
```

```{r}
geocode_OSM("949 wealthy Street SE Grand Rapids")
```


```{r}
test <- rev_geocode_OSM(x = -85.67486	, y = 42.94856)[[1]]["postcode"]

test
```

```{r}
#t <-  map2(crash_data$X, crash_data$Y, rev_geocode_OSM)

#head(t)
```


```{r}
crash_data_zip <- crash_data %>% 
  filter(YEAR == 2017, MONTH == "June", DAYOFMONTH == 1) %>% 
  select(X,Y) %>% 
  rowwise() %>% 
  mutate(zip_code = unlist(rev_geocode_OSM(x = X, y = Y)[[1]]["postcode"]))

head(crash_data_zip)
```

```{r}
location_gr <- getbb("Grand Rapids") %>% 
  opq()

roads_gr <- location_gr %>%
  add_osm_feature(key = "highway", value = c("motorway", "trunk", "primary", "secondary", "tertiary")) %>%
  osmdata_sf()

water_gr <- location_gr %>% 
  add_osm_feature(key = "waterway", value = c("river")) %>% 
  osmdata_sf()

boundary_gr <- location_gr %>% 
  add_osm_feature(key = "boundary", value = "administrative") %>%
  #add_osm_feature(key = "admin_level", value = "8") %>% 
  #add_osm_feature(key = "border_type", value = "city") %>% 
  #add_osm_feature(key = "place", value = "city") %>% 
  add_osm_feature(key = "name", value = "Grand Rapids") %>% 
  osmdata_sf()

#zip_gr <- location_gr %>% 
 # add_osm_feature(key = "boundary", value = "postal_code") %>% 
 # add_osm_feature(key = "postal_code", value = "49506") %>% 
  #osmdata_sf()

```

```{r}
ggplot()+
         geom_sf(data = roads_gr$osm_lines, size = .5, alpha = .5, color = 'black') +
         geom_sf(data = water_gr$osm_lines, size = 1, alpha = .3, color = 'steelblue') +
         geom_sf(data = boundary_gr$osm_lines, size = 1, alpha = .5, color = "orange") +
         #geom_sf(data = zip_gr$osm_lines, size = .5, color = "red") +
         coord_sf(xlim = c(-85.57, -85.75), ylim = c(42.88, 43.03))
```


```{r}
zip_gr <- location_gr %>% 
  add_osm_feature(key = "boundary", value = "") %>% 
  add_osm_feature(key = "boundary", value = "postal_code") %>% 
  osmdata_sf()

zip_gr
```

```{r}
boundary_gr <- location_gr %>% 
  add_osm_feature(key = "boundary", value = "administrative") %>%
  #add_osm_feature(key = "admin_level", value = "8") %>% 
  #add_osm_feature(key = "border_type", value = "city") %>% 
  #add_osm_feature(key = "place", value = "city") %>% 
  add_osm_feature(key = "name", value = "Grand Rapids") %>% 
  osmdata_sf()

boundary_gr
```


```{r}
crash_data$HITANDRUN <- if_else(crash_data$HITANDRUN == "Yes", 1, 0)

logistic <- glm(HITANDRUN ~ HOUR, data = crash_data, family = "binomial")

summary(logistic)
```

```{r}
rr_crossing_data <- read_csv(here::here("data", "Crossing_Inventory_Data_-_Current.csv"))

head(rr_crossing_data)
```

```{r}
rr_crossing_data_gr <- rr_crossing_data %>% 
  filter(`State Name` == "MICHIGAN", `City Name` == "GRAND RAPIDS")

rr_crossing_data_gr
```

```{r}
rr_crash <- crash_data %>% 
  rowwise() %>% 
  filter(1 %in% in_radius(X,Y))

rr_crash
```


```{r}
ggplot()+
         geom_sf(data = roads_gr$osm_lines, size = .5, alpha = .5, color = 'black') +
         geom_sf(data = water_gr$osm_lines, size = 1, alpha = .3, color = 'steelblue') +
         geom_sf(data = boundary_gr$osm_lines, size = 1, alpha = .5, color = "orange") +
         geom_point(data = rr_crossing_data_gr, mapping = aes(x = Longitude, y = Latitude), color = "red") +
         geom_point(data = rr_crash, mapping = aes(x = X, y = Y), color = "blue") +
         geom_circle(data = rr_crossing_data_gr, mapping = aes(x0 = Longitude, y0 = Latitude, r = .0005), color = "red") +
         coord_sf(xlim = c(-85.57, -85.75), ylim = c(42.88, 43.03))
         #coord_sf(xlim = c(-85.676, -85.685), ylim = c(42.96, 42.9665))
```

```{r}
in_radius <- function(x, y) {
  if_else(((x - rr_crossing_data_gr$Longitude) ^ 2 + (y - rr_crossing_data_gr$Latitude) ^ 2) <= .0005 ^ 2, 1, 0)
}
```

```{r}
in_radius(-85.63890,42.92721)
```

```{r}
rr_crash %>% 
  ggplot() +
  geom_bar(mapping = aes(x = CRASHTYPE, color = CRASHTYPE))
```

```{r}
crash_data %>% 
  ggplot() +
  geom_bar(mapping = aes(x = CRASHTYPE, color = CRASHTYPE))
```

The number of accidents in our radius gives us an idea of exposure

```{r}
rr_crossing_data_gr %>% 
  filter(`Count Roadway Gate Arms` > 0)
```

