---
title: "Train Accident Findings Report"
author: "Luke Cadagin"
date: "11/24/2021"
output:
  github_document: default
  html_document: default
  pdf_document: default
---

# Report of Grand Rapids Train Accident Analysis

## Set Up

Load necessary packages and set max numerical digits displayed in tibble to 8 (necessary for long/lat coordinates):

```{r message = FALSE}
library(tidyverse)
library(sf)
library(osmdata)
library(ggpubr)
library(ggforce)
options(digits = 8)
```

Specify size of all graphs in Knitted Documents:

```{r}
knitr::opts_chunk$set(echo = TRUE, fig.width = 16, fig.height = 8)
```


Upload Grand Rapids Crash Dataset (2008 - 2017):

```{r message = FALSE}
crash_data <- read_csv(here::here("data", "CGR_Crash_Data.csv"))
head(crash_data[1:6])
```

Configure features necessary for Grand Rapids map (using openstreetmap api)

```{r}
location_gr <- getbb("Grand Rapids") %>% 
    opq()

major_roads_gr <- location_gr %>%
    add_osm_feature(key = "highway", value = c("motorway", "trunk", "primary", "secondary", "tertiary")) %>% 
    osmdata_sf()

#minor_roads_gr <- location_gr %>%
    #add_osm_feature(key = "highway", value = c("unclassified", "residential")) %>%
    #osmdata_sf()

water_gr <- location_gr %>% 
    add_osm_feature(key = "waterway", value = c("river")) %>% 
    osmdata_sf()

boundary_gr <- location_gr %>% 
    add_osm_feature(key = "boundary", value = "administrative") %>%
    add_osm_feature(key = "name", value = "Grand Rapids") %>% 
    osmdata_sf()
```

## Initial Exploration of Direct Train Accidents

I am interested in studying the impact gates at a railroad crossing have on the number of accidents associated with trains.

Let's start by visualizing the number of Grand Rapids crashes associated with a train from 2008 - 2017.

```{r}
crash_data_train <- crash_data %>% 
  filter(TRAIN == "Yes")

crash_data_train %>% 
  select(`Longitude` = X, `Latitude` = Y, CRASHDATE, TRAIN, `Principal Road` = PRNAME)
```

```{r warning = FALSE}
  ggplot()+
            geom_sf(data = major_roads_gr$osm_lines, size = .6, alpha = .6, color = 'black') +
            #geom_sf(data = minor_roads_gr$osm_lines, size = .3, alpha = .3, color = 'black') +
            geom_sf(data = water_gr$osm_lines, size = 1, alpha = .4, color = 'steelblue') +
            geom_sf(data = boundary_gr$osm_lines, size = 1, alpha = .6, color = "orange") +
            geom_point(data = crash_data_train, mapping = aes(x = X, y = Y), color = "blue") +
            coord_sf(xlim = c(-85.57, -85.75), ylim = c(42.88, 43.03)) +
            labs(title = "Grand Rapids City Limits", x = "Longitude", y = "Latitude") +
            font("title", size = 20, color = "blue", face = "bold") +
            font("x", size = 16) +
            font("y", size = 16)
```

As you can see, from 2008 to 2017 there were only three accidents that occurred in Grand Rapids involving a train (In 2017, 2014, and 2008).

## Pulling In Grand Rapids Train Crossings:

Although we have a very limited number of accidents directly involving a train, this is not a dead-end for our analysis.

Next let's upload a a dataset from Transportation.gov (https://data.transportation.gov/Railroads/Crossing-Inventory-Data-Current/m2f8-22s6) that provides information about every railroad crossing in the USA:

```{r message = FALSE, warning = FALSE}
rr_crossing_data <- read_csv(here::here("data", "Crossing_Inventory_Data_-_Current.csv"))

head(rr_crossing_data[1:6])
```

Now we filter this data for only railroad crossing within the Grand Rapids city limits (Note that there are three crossing located in the city center where Latitude = 42.96336 that I removed as it seems like the long/lang for these were placeholders):

```{r}
rr_crossing_data_gr <- rr_crossing_data %>% 
  filter(`State Name` == "MICHIGAN", `City Name` == "GRAND RAPIDS", `Intersecting Roadway` == "Yes", Latitude < 43, Latitude != 42.96336)#, Latitude > 42.88, Longitude > -85.73, Longitude < -85.59, (Latitude > 42.92 | Longitude > -85.67))

head(rr_crossing_data_gr[1:6])
```

Let's Make Sure that there are no duplicates in the data:

```{r}
nrow(distinct(rr_crossing_data_gr, Latitude))
```

```{r}
nrow(rr_crossing_data_gr)
```

We can visualize this data in red on our Grand Rapid's Map:

```{r fig.width = 16, fig.height = 8, warning = FALSE}
  ggplot()+
            geom_sf(data = major_roads_gr$osm_lines, size = .6, alpha = .6, color = 'black') +
            #geom_sf(data = minor_roads_gr$osm_lines, size = .3, alpha = .3, color = 'black') +
            geom_sf(data = water_gr$osm_lines, size = 1, alpha = .4, color = 'steelblue') +
            geom_sf(data = boundary_gr$osm_lines, size = 1, alpha = .6, color = "orange") +
            geom_point(data = rr_crossing_data_gr, mapping = aes(x = Longitude, y = Latitude), color = "red") +
            coord_sf(xlim = c(-85.57, -85.75), ylim = c(42.88, 43.03)) +
            labs(title = "Grand Rapids City Limits", x = "Longitude", y = "Latitude") +
            font("title", size = 20, color = "blue", face = "bold") +
            font("x", size = 16) +
            font("y", size = 16)
```

## Exploring Accidents in the Vacinity of Railroad Crossings:

We now would like to know how many crashes fall within a .0005 (longitudinal units) radius of each railroad crossing.

To do so, we first write a function called in_radius() that detects if a longitude/latitude coordinate is located within a .0005 radius of all railroad crossings:

```{r}
in_radius <- function(x1, y1, x2, y2) {
  if_else(((x1 - x2) ^ 2 + (y1 - y2) ^ 2) <= .0005 ^ 2, 1, 0)
}
```

Next we create a new accident subset named near_rr_crash that only contains crashes that occurred within a .0005 radius of a railroad crossing:

```{r}
near_rr_crash <- crash_data %>% 
  rowwise() %>% 
  filter(1 %in% in_radius(X, Y, rr_crossing_data_gr$Longitude, rr_crossing_data_gr$Latitude))

head(near_rr_crash[1:6])
```

We plot this data on our grand rapids map to visualize the result:

```{r fig.width = 16, fig.height = 8, warning = FALSE}
  ggplot()+
            geom_sf(data = major_roads_gr$osm_lines, size = .6, alpha = .6, color = 'black') +
            #geom_sf(data = minor_roads_gr$osm_lines, size = .3, alpha = .3, color = 'black') +
            geom_sf(data = water_gr$osm_lines, size = 1, alpha = .4, color = 'steelblue') +
            geom_sf(data = boundary_gr$osm_lines, size = 1, alpha = .6, color = "orange") +
            geom_point(data = near_rr_crash, mapping = aes(x = X, y = Y), color = "blue") +
            geom_point(data = rr_crossing_data_gr, mapping = aes(x = Longitude, y = Latitude), color = "red") +
            geom_circle(data = rr_crossing_data_gr, mapping = aes(x0 = Longitude, y0 = Latitude, r = .0005), color = "red") +
            coord_sf(xlim = c(-85.57, -85.75), ylim = c(42.88, 43.03)) +
            labs(title = "Grand Rapids City Limits", x = "Longitude", y = "Latitude") +
            font("title", size = 20, color = "blue", face = "bold") +
            font("x", size = 16) +
            font("y", size = 16)
```

Below is a zoomed-in portion of the map for better detail (located around the GVSU Pew Campus):

```{r fig.width = 16, fig.height = 8, warning = FALSE}
  ggplot()+
            geom_sf(data = major_roads_gr$osm_lines, size = .6, alpha = .6, color = 'black') +
            #geom_sf(data = minor_roads_gr$osm_lines, size = .3, alpha = .3, color = 'black') +
            geom_sf(data = water_gr$osm_lines, size = 1, alpha = .4, color = 'steelblue') +
            geom_sf(data = boundary_gr$osm_lines, size = 1, alpha = .6, color = "orange") +
            geom_point(data = near_rr_crash, mapping = aes(x = X, y = Y), color = "blue") +
            geom_point(data = rr_crossing_data_gr, mapping = aes(x = Longitude, y = Latitude), color = "red") +
            geom_circle(data = rr_crossing_data_gr, mapping = aes(x0 = Longitude, y0 = Latitude, r = .0005), color = "red") +
            coord_sf(xlim = c(-85.676, -85.685), ylim = c(42.96, 42.9665)) +
            labs(title = "Grand Rapids City Limits", x = "Longitude", y = "Latitude") +
            font("title", size = 20, color = "blue", face = "bold") +
            font("x", size = 16) +
            font("y", size = 16)
```

## Summarizing and Analyzing Results:

Now we create a tibble that sums the number of crashes that fall within a .0005 radius of each crossing along with the number of Gate arms at that crossing and the total number of trains that pass through during the day and night:

```{r}
exposure <- rr_crossing_data_gr %>% 
  rowwise() %>% 
  mutate(Radius_Count = sum(in_radius(Longitude, Latitude, near_rr_crash$X, near_rr_crash$Y))) %>% 
  select(Longitude, Latitude, Street, "Gate_Arm_N" = `Count Roadway Gate Arms`, Radius_Count, "Day_Trains_N" = `Total Daylight Thru Trains`, "Night_Trains_N" = `Total Nighttime Thru Trains`) %>% 
  arrange(desc(Radius_Count)) %>% 
  ungroup()

head(exposure,10)
```

We can use the number of accidents that occur within .0005 longitudinal units from a crossing (Radius_Count) to represent/correspond to traffic exposure.  We are deriving latent information from this variable to determine if additional safety protocols should be put in place at any particular train crossing.

The tibble shows us the crossings with the highest number of crashes that occurred near-by (Radius_Count).  We see that out of the top 10 crossings with the highest traffic exposure, only 3 have gate arms.

Let's take a look at these top 10 traffic exposure crossings on the Grand Rapids Map:

```{r}
exposure_top_10 <- exposure %>% 
  slice_head(n = 10)
```

```{r warning = FALSE}
  ggplot()+
            geom_sf(data = major_roads_gr$osm_lines, size = .6, alpha = .6, color = 'black') +
            #geom_sf(data = minor_roads_gr$osm_lines, size = .3, alpha = .3, color = 'black') +
            geom_sf(data = water_gr$osm_lines, size = 1, alpha = .4, color = 'steelblue') +
            geom_sf(data = boundary_gr$osm_lines, size = 1, alpha = .6, color = "orange") +
            geom_point(data = exposure_top_10, mapping = aes(x = Longitude, y = Latitude), color = "red") +
            coord_sf(xlim = c(-85.57, -85.75), ylim = c(42.88, 43.03)) +
            labs(title = "Grand Rapids City Limits", x = "Longitude", y = "Latitude") +
            font("title", size = 20, color = "blue", face = "bold") +
            font("x", size = 16) +
            font("y", size = 16)
```

We can see these crossings are scattered around the city.  Notice that two of these crossings are located on the GVSU Pew Campus:

```{r warning = FALSE, message =FALSE}
  ggplot()+
            geom_sf(data = major_roads_gr$osm_lines, size = .6, alpha = .6, color = 'black') +
            #geom_sf(data = minor_roads_gr$osm_lines, size = .3, alpha = .3, color = 'black') +
            geom_sf(data = water_gr$osm_lines, size = 1, alpha = .4, color = 'steelblue') +
            geom_sf(data = boundary_gr$osm_lines, size = 1, alpha = .6, color = "orange") +
            geom_point(data = exposure_top_10, mapping = aes(x = Longitude, y = Latitude), color = "red") +
            #geom_circle(data = exposure_top_10, mapping = aes(x0 = Longitude, y0 = Latitude, r = .0005), color = "red") +
            coord_sf(xlim = c(-85.676, -85.685), ylim = c(42.96, 42.9665)) +
            labs(title = "Grand Rapids City Limits", x = "Longitude", y = "Latitude") +
            geom_text(data = exposure_top_10, mapping = aes(Longitude, Latitude, label = Street), nudge_y = -.0004, color = "red") +
            font("title", size = 20, color = "blue", face = "bold") +
            font("x", size = 16) +
            font("y", size = 16)
```

We can see that the crossing at "LAKE MICHIGAN DRIVE NW" (-85.681461,42966083) does have gate arms while the crossing at "FULTON STREET" (-85.681619,42.963392	) does not have gate arms.  Also note that the total Daylight (1) and Nighttime (2) trains for these crossings are the same while the Radius_Count for the "FULTON STREET" crossing (45) is greater than the "LAKE MICHIGAN DRIVE NW" crossing (40):

```{r}
exposure_top_10 %>% 
  filter(Street %in% c("LAKE MICHIGAN DRIVE NW", "FULTON STREET"))
```

I am not sure what metrics are used by the city of Grand Rapids to determine if a railroad crossing arm is necessary, but there appears to be a degree of inconsistency.

Note that among all 74 railroad crossings in Grand Rapids, 26 have at least one gate arm.

```{r}
exposure %>% 
  filter(Gate_Arm_N > 0) %>% 
  nrow()
```

Of these 26 crossings, 22 have at least one Daylight Train OR one Nighttime train pass through.  This is about 85% (22/26):

```{r}
exposure %>% 
  filter(Gate_Arm_N > 0, (Day_Trains_N > 0 | Night_Trains_N > 0)) %>% 
  nrow()
```

With this in mind, let us take a second look at the top 10 traffic exposure crossings:

```{r}
exposure_top_10
```

As we noted earlier, 7 of these crossings do not have gate arms:

```{r}
exposure_top_10 %>% 
  filter(Gate_Arm_N == 0)
```

Of these 7, 5 have at least one daylight OR nighttime train go through each day:

```{r}
exposure_top_10 %>% 
  filter(Gate_Arm_N == 0, (Day_Trains_N > 0 | Night_Trains_N > 0))
```

I would recommend that the city of Grand Rapids takes a second look at these crossings to determine if a gate arm or additional traffic precautions are necessary.

# Exploratory Application

To supplement this analysis, I have put together a Shiny Application for the city of Grand Rapids to explore the crash dataset on their own.  This application can be accessed using the address below:

https://lcadagin.shinyapps.io/interactive_crash_plotting/

